MLH1 Results Outline

### 2020-02-07 doc/Results.rmd
female specific analysis patterns

## Male Specific Polymorphism

@@

-- what's the driving question for this section --

@@

@@

@@

@@

bhh11 <- lm(mean_co ~ subsp * strain, data= fomo.male_mouse.av)
#bhh11$coefficients

booo <- summary(bhh11)

lm.male.strain <- lm(mean_co ~ strain, data= fomo.male_mouse.av)
ll <- summary(lm.male.strain)

#even when all the other male strains are added -- the same result; PWD, MSM and SKIVE strain effects are significant.
```

@@

@@

@@

@@

#### SC_HetC_M1

@@

# Q2 Analysis Predictions, Male Polymorphism and (High vs Low Rec strains)

@@

#these 2 below are LR tests
drop1(lmer.IFD.PER_MM.M1, test="Chisq" )
anova(lmer.IFD.PER_MM.M1)

exactRLRT(lmer.IFD.PER_MM.M1)
as.data.frame(round(summary(lmer.IFD.PER_MM.M1)$coefficients, 6))

@@

#these 2 below are LR tests
drop1(lmer.IFD.RAW_MM.M1, test="Chisq" ) #other options for test don't work
anova(lmer.IFD.RAW_MM.M1) # also F values

exactRLRT(lmer.IFD.RAW_MM.M1)#random strain not sig
as.data.frame(round(summary(lmer.IFD.RAW_MM.M1)$coefficients, 6))#these Dont return p values for the fixed effects
```

@@

@@

#table(WSB.Bivalents.DF$sex, WSB.Bivalents.DF$hand.foci.count)#the 2CO's will be used

@@

```{r LEW.IFD, echo=FALSE, warning=FALSE}
#make the dataset
LEW.Bivalents.DF <- Curated_BivData[Curated_BivData$strain == "LEW",]
#table(LEW.Bivalents.DF$sex, LEW.Bivalents.DF$hand.foci.count)

@@

lew.ploty_IFD_raw <- ggplot(LEW.Bivalents.DF[LEW.Bivalents.DF$hand.foci.count == 2,],  aes(y=IFD1_ABS, x = sex))+
                          geom_jitter()+ggtitle("Raw IFD, LEW")+geom_boxplot(aes(x=sex))
lew.ploty_IFD_raw

@@

LEW.IFDnorm_scatter.plot <- ggplot(LEW.Bivalents.DF[LEW.Bivalents.DF$hand.foci.count == 2,],  aes(y=IFD1_PER, x = chromosomeLength, color = sex))+                          geom_jitter()+ggtitle("LEW \n Normalized IFD and SC Length")+ylim(c(0,1.1))

@@

                  n_2CO =  sum(hand.foci.count == 2, na.rm = TRUE) #this is the right syntax
)
#can I plot these stats

@@

top.male <-  G.Bivalents.DF %>% filter(sex== "male") %>% filter(IFD1_PER <= .3) #this makes DF
bottom.male <-  G.Bivalents.DF %>% filter(sex== "male")
ratio.male <- 18/1031 #2%

top.female <-  G.Bivalents.DF %>% filter(sex== "female") %>% filter(IFD1_PER <= .3)
bottom.female <-  G.Bivalents.DF %>% filter(sex== "female")
ratio.female <- 36/766  #5%

@@

PWD.ploty_IFD_PER <- ggplot(PWD.Bivalents.DF[PWD.Bivalents.DF$hand.foci.count == 2,],  aes(y=IFD1_PER, x = sex))+
                          geom_jitter()+ggtitle("Percent IFD, PWD")+geom_boxplot(aes(x=sex))

PWD.IFDnorm_scatter.plot <- ggplot(PWD.Bivalents.DF[PWD.Bivalents.DF$hand.foci.count == 2,],  aes(y=IFD1_PER, x = chromosomeLength, color = sex))+ geom_jitter()+ggtitle("PWD \n Normalized IFD and SC Length")+ylim(c(0,1.1))

@@

#make a table of the means and n ect
PWD.table <-  ddply(PWD.Bivalents.DF, c("sex"), summarise,

@@

```{r KAZ.IFD, echo=FALSE, warning=FALSE}
#make the dataset
KAZ.Bivalents.DF <- Curated_BivData[Curated_BivData$strain == "KAZ",]

@@

KAZ.IFDnorm_scatter.plot <- ggplot(KAZ.Bivalents.DF[KAZ.Bivalents.DF$hand.foci.count == 2,],  aes(y=IFD1_PER, x = chromosomeLength, color = sex))+ geom_jitter()+ggtitle("KAZ \n Normalized IFD and SC Length")+ylim(c(0,1.1))

@@

#make a table of the means and n ect
KAZ.table <-  ddply(KAZ.Bivalents.DF, c("sex"), summarise,

@@

```{r MOLF.IFD, eval=FALSE, echo=FALSE}
#make the dataset
MOLF.Bivalents.DF <- Curated_BivData[Curated_BivData$strain == "MOLF",]
table(MOLF.Bivalents.DF$sex, MOLF.Bivalents.DF$hand.foci.count)

@@

@@

        #           mean.SC_0CO = mean(chromosomeLength[which(hand.foci.count == 0)], na.rm = TRUE),
         #         n_0CO =  sum((hand.foci.count == 0), na.rm = TRUE ),
                 # CO2 = sum(hand.foci.count == 2, na.rm=TRUE),

      #             mean.SC_1CO = mean(chromosomeLength[which(hand.foci.count == 1)], na.rm = TRUE),
      #              n_1CO = sum((hand.foci.count == 1), na.rm = TRUE),

      #             mean.SC_2CO = mean(chromosomeLength[which(hand.foci.count == 2)], na.rm = TRUE),
      #             n_2CO = sum((hand.foci.count == 2), na.rm = TRUE),

      #            mean.SC_3CO = mean(chromosomeLength[which(hand.foci.count == 3)], na.rm = TRUE),
      #            n_3CO = sum((hand.foci.count == 3), na.rm = TRUE)

@@

```{r Q2.SC.ploy.mal, echo=FALSE}
# 1 way anova, is at least 1 strain different from the rest?
# given a lm, how sig is the strain effect

@@

>>>>

# Female Specific Polymorphism

I should run the same questions for the female data --

```{r female.show.plots, echo=FALSE, fig.width = 4, fig.height=2.5}

female_mouse.av <- subset(AP_mouse_table_w.Ages, select= -c(sex.y,sex.x,
                            subsp.y,subsp.x,
                    category.y,category.x, strain.x,
                    strain.y, mouse.1  )  )

female_mouse.av <- add_sex(female_mouse.av)
female_mouse.av <- add_strain(female_mouse.av)
female_mouse.av <- add_subsp(female_mouse.av)
female_mouse.av <- add_species(female_mouse.av)

female_mouse.avs_for.analysis <-  female_mouse.av %>% filter(sex== "female") %>% filter(species == "M.musculus")

female_mouse.avs_for.analysis$subsp <- factor(female_mouse.avs_for.analysis$subsp, ordered = FALSE)
female_mouse.avs_for.analysis$sex <- factor(female_mouse.avs_for.analysis$sex, ordered = FALSE)
female_mouse.avs_for.analysis$strain <- factor(female_mouse.avs_for.analysis$strain, ordered = FALSE)
female_mouse.avs_for.analysis$species <- factor(female_mouse.avs_for.analysis$species, ordered = FALSE)

table( droplevels(female_mouse.avs_for.analysis$strain) )
#there are cast strains too

Musc.female_mouse.av <-  female_mouse.avs_for.analysis %>% filter(sex== "female") %>% filter(subsp == "Musc")

musc.female.plot <- ggplot(Musc.female_mouse.av, aes(color= strain))+geom_jitter(aes(x = as.factor(strain), y = mean_co, color= strain, alpha=0.5), width = .2 )+ ggtitle("Musc female of MLH1 distributions")+scale_color_manual(values=colors_of_strains)+ylim(c(20,34))
musc.male.plot

#Mol
Mol.female_mouse.av <-  female_mouse.avs_for.analysis %>% filter(sex== "female") %>% filter(subsp == "Mol")

Mol.female.plot <- ggplot(Mol.female_mouse.av, aes(color= strain))+geom_jitter(aes(x = as.factor(strain), y = mean_co, color= strain, alpha=0.5), width = .2 )+ ggtitle("Mol female of MLH1 distributions")+scale_color_manual(values=colors_of_strains)+ylim(c(20,33))
Mol.female.plot

#Dom version 
Dom.female_mouse.av <-  female_mouse.avs_for.analysis %>% filter(sex== "female") %>% filter(subsp == "Dom")

Dom.female.plot <- ggplot(Dom.female_mouse.av, aes(color= strain))+geom_jitter(aes(x = as.factor(strain), y = mean_co, color= strain, alpha=0.5), width = .2 )+ ggtitle("Dom female of MLH1 distributions")+scale_color_manual(values=colors_of_strains)+ylim(c(20,33))
Dom.female.plot
#Gough female is higher

#variance
musc.female.VAR.plot <- ggplot(Musc.female_mouse.av, aes(color= strain))+geom_jitter(aes(x = as.factor(strain), y =var, color= strain, alpha=0.5), width = .2 )+ ggtitle("Musc female of within mouse MLH1 distributions")+scale_color_manual(values=colors_of_strains)
musc.female.VAR.plot

Mol.female.plot <- ggplot(Mol.female_mouse.av, aes(color= strain))+geom_jitter(aes(x = as.factor(strain), y = var, color= strain, alpha=0.5), width = .2 )+ ggtitle("Mol female of within mouse MLH1 distributions")+scale_color_manual(values=colors_of_strains)
Mol.female.plot

Dom.female.plot <- ggplot(Dom.female_mouse.av, aes(color= strain))+geom_jitter(aes(x = as.factor(strain), y = var, color= strain, alpha=0.5), width = .2 )+ ggtitle("Dom feale ofwithin mouse  MLH1 distributions")+scale_color_manual(values=colors_of_strains)
Dom.female.plot

```{r Female.specific.test, echo=FALSE}

lm.female.strain.subsp <- lm(mean_co ~ subsp * strain, data= female_mouse.avs_for.analysis)
summry.female.strain.effects <- summary(lm.female.strain.subsp)
#strainG, 1.06e-05 ***
#strainLEW  0.0273 *

lm.female.strain <- lm(mean_co ~ strain, data= female_mouse.avs_for.analysis)
lm.female.strain_sum <- summary(lm.female.strain)
#G  1.06e-05 ***
#LEW   0.0273 *
#PWD  0.0810 .
#MSM  3.02e-05 ***

#how much greater is G female
G_female <-  female_mouse.avs_for.analysis %>% filter(strain == "G")
mean.G.female <- mean(G_female$mean_co)

sans_G_female <- female_mouse.avs_for.analysis %>% filter(strain != "G")
mean.all.other <- mean(sans_G_female$mean_co)

## within mouse variance models 
lmVAR.female.strain.subsp <- lm(var ~ subsp * strain, data= female_mouse.avs_for.analysis)
summry.female.strain.VAReffects <- summary(lmVAR.female.strain.subsp)

lmVAR.female.strain <- lm(var ~ strain, data= female_mouse.avs_for.analysis)
lmVAR.female.strain_sum <- summary(lmVAR.female.strain)

#Q12
Q12_female_table <-  Q12_mouse_table_w.Ages %>% filter(sex == "female")

lm.m1.Q12_female <- lm(var ~ subsp * strain, data= Q12_female_table)
lm.m1.Q12_female_sum <- summary(lm.m1.Q12_female)
#LEW not sig

lm.m2.Q12_female <- lm(var ~ strain, data= Q12_female_table)
lm.m2.Q12VAR.female_sum <- summary(lmVAR.female.strain)
#LEW sig
```

# Male Specific Polymorphism

@@

@@

just.male_mouse.av <-  male_mouse.av %>% filter(sex== "male") %>% filter(species == "M.musculus")

just.male_mouse.av$sex <- factor(just.male_mouse.av$sex, ordered = FALSE)
just.male_mouse.av$strain <- factor(just.male_mouse.av$strain, ordered = FALSE)
just.male_mouse.av$subsp <- factor(just.male_mouse.av$subsp, ordered = FALSE)
just.male_mouse.av$species <- factor(just.male_mouse.av$species, ordered = FALSE)

#table( droplevels(just.male_mouse.av$strain) )
#there are cast strains too

Musc.male_mouse.av <-  just.male_mouse.av %>% filter(sex== "male") %>% filter(subsp == "Musc")

@@

#make the Mol and
Mol.male_mouse.av <-  just.male_mouse.av %>% filter(sex== "male") %>% filter(subsp == "Mol")

@@

#Dom version 
Dom.male_mouse.av <-  just.male_mouse.av %>% filter(sex== "male") %>% filter(subsp == "Dom")

@@

lm.male.strain.subsp <- lm(mean_co ~ subsp * strain, data= just.male_mouse.av)
summry.male.strain.effects <- summary(lm.male.strain.subsp)

lm.male.strain <- lm(mean_co ~ strain, data= just.male_mouse.av)
lm.male.strain_sum <- summary(lm.male.strain)

#even when all the other male strains are added -- the same result; PWD, MSM and SKIVE strain effects are significant.

#how much greater are the High Rec gwRRs
just.male_mouse.av$Rec.group <- ifelse(grepl("PWD", just.male_mouse.av$strain), 1, 
                              ifelse(grepl("MSM", just.male_mouse.av$strain), 1,
                                ifelse(grepl("SKIVE", just.male_mouse.av$strain), 1, 0)))

sans_LewRec_male <- just.male_mouse.av %>% filter(Rec.group == 0)
#mean(sans_LewRec_male$mean_co)

PWD_male <-  just.male_mouse.av %>% filter(strain == "PWD")
#mean(PWD_male$mean_co) / mean(sans_LewRec_male$mean_co)

MSM_male <- just.male_mouse.av %>% filter(strain == "MSM")
#mean(MSM_male$mean_co) / mean(sans_LewRec_male$mean_co)

SKIVE_male <-   just.male_mouse.av %>% filter(strain == "SKIVE")
#mean(SKIVE_male$mean_co) /  
  mean(sans_LewRec_male$mean_co)

## within mouse variance models 
lmVAR.male.strain.subsp <- lm(var ~ subsp * strain, data= just.male_mouse.av)
summry.male.strain.VAReffects <- summary(lmVAR.male.strain.subsp)
#all p values kinda high, MSM and TOM are slightly sig

lmVAR.male.strain <- lm(var ~ strain, data= just.male_mouse.av)
lmVAR.male.strain_sum <- summary(lmVAR.male.strain)
#non sig

lmcv.male.strain <- lm(cV ~ strain, data= just.male_mouse.av)
lmcv.male.strain_sum <- summary(lmcv.male.strain)
#HMI sig for cv model -- in the other model, CAST subsp and TOM strain are slightly significant

#Q12
Q12_male_table <-  Q12_mouse_table_w.Ages %>% filter(sex == "male")

lm.m1.Q12_male <- lm(var ~ subsp * strain, data= Q12_male_table)
lm.m1.Q12_male_sum <- summary(lm.m1.Q12_male)
#NS

lm.m2.Q12_male <- lm(var ~ strain, data= Q12_male_table)
lm.m2.Q12VAR.male_sum <- summary(lm.m2.Q12_male)
#NS

@@

TODO -- try running a chi square test on the proportions.

@@

As alex said, be careful about mixed models and interperting the random effect (radnom variance) -- assigning strain as random effect is slightly problamatic as i can give it a biological interpertation (it's due to strain -- genetics of each of the strains).

@@

## Q2 Analysis Predictions, Male Polymorphism and (High vs Low Rec strains)

@@

#### Q1.SC_HetC_M1

@@

@@

#these 2 below are LR tests
#drop1(lmer.IFD.PER_MM.M1, test="Chisq" )
#anova(lmer.IFD.PER_MM.M1)

#exactRLRT(lmer.IFD.PER_MM.M1)
#as.data.frame(round(summary(lmer.IFD.PER_MM.M1)$coefficients, 6))

@@

#these 2 below are LR tests
#drop1(lmer.IFD.RAW_MM.M1, test="Chisq" ) #other options for test don't work
#anova(lmer.IFD.RAW_MM.M1) # also F values

#exactRLRT(lmer.IFD.RAW_MM.M1)#random strain not sig
#as.data.frame(round(summary(lmer.IFD.RAW_MM.M1)$coefficients, 6))#these Dont return p values for the fixed effects
```

@@

```{r Comp_Ratios, echo=FALSE}

#calq ratios of belo 30

top.male <-  G.Bivalents.DF %>% filter(sex== "male") %>%  filter(hand.foci.count== 2) %>%filter(IFD1_PER <= .3)

bottom.male <-  G.Bivalents.DF %>% filter(sex== "male") %>%  filter(hand.foci.count== 2)

ratio.male <- length(top.male$Obj.ID)/ length(bottom.male$Obj.ID)

top.female <-  G.Bivalents.DF %>% filter(sex== "female") %>%  filter(hand.foci.count== 2) %>% filter(IFD1_PER <= .3)

bottom.female <-  G.Bivalents.DF %>% filter(sex== "female")%>%  filter(hand.foci.count== 2)

ratio.female <- length(top.female$Obj.ID)/ length(bottom.female$Obj.ID)

#what is the fastest way to calq all these ratios?  also what's the fastst way to get them into a DF / table?

#put all the DFs in a list, for each DF in the list -- apply the filters for the nom and denominator

DF.strain.list <- list(
  WSB.Bivalents.DF, G.Bivalents.DF, LEW.Bivalents.DF, 
  PWD.Bivalents.DF, KAZ.Bivalents.DF, SKIVE.Bivalents.DF,
  MSM.Bivalents.DF, MOLF.Bivalents.DF
  )

entr_DF <- data.frame(strain=c("WSB", "G", "LEW", "PWD", "KAZ", "SKIVE", "MSM", "MOLF"), 
        total.2CO_female=c(0,0,0,0,0,0,0,0), total.2CO_male=c(0,0,0,0,0,0,0,0), below30_male = c(0,0,0,0,0,0,0,0), below30_female = c(0,0,0,0,0,0,0,0),     
          female.ratio = c(0,0,0,0,0,0,0,0), male.ratio = c(0,0,0,0,0,0,0,0))

i = 1
for(x in (DF.strain.list) ){

  #print(head())

  #these filters won't apply to the indexed DF
  top.male =  x %>% filter(sex== "male") %>%  filter(hand.foci.count== 2) %>%filter(IFD1_PER <= .3)

  bottom.male <-  x %>% filter(sex== "male") %>%  filter(hand.foci.count== 2)

  entr_DF$total.2CO_male[i] <- length(bottom.male$Obj.ID)
  entr_DF$below30_male[i] <- length(top.male$Obj.ID)
  entr_DF$male.ratio[i] <- length(top.male$Obj.ID)/ length(bottom.male$Obj.ID)

  top.female <-  x %>% filter(sex== "female") %>%  filter(hand.foci.count== 2) %>% filter(IFD1_PER <= .3)

  bottom.female <-  x %>% filter(sex== "female")%>%  filter(hand.foci.count== 2)

  entr_DF$total.2CO_female[i] <- length(bottom.female$Obj.ID)
  entr_DF$below30_female[i] <- length(top.female$Obj.ID)
  entr_DF$female.ratio[i] <- length(top.female$Obj.ID)/ length(bottom.female$Obj.ID)

  i = i+1

  }
entr_DF

entr_DF$sex.dif <- entr_DF$female.ratio - entr_DF$male.ratio

Above is a table of the proportion of 2Co bivalents which have a norm IFD below 30%, For all strains but KAZ, the females have a greater proportion of these shorted IFD values.

@@

#table(WSB.Bivalents.DF$sex, WSB.Bivalents.DF$hand.foci.count)#the 2CO's will be used
WSB.Bivalents.DF <- Curated_BivData[Curated_BivData$strain == "WSB",]

@@

```{r LEW.IFD, echo=FALSE, warning=FALSE}
#make the dataset
LEW.Bivalents.DF <- Curated_BivData[Curated_BivData$strain == "LEW",]

LEW.Bivalents.DF_IFD <- LEW.Bivalents.DF[(!is.na(LEW.Bivalents.DF$sex)) & (LEW.Bivalents.DF$hand.foci.count == 2),]

LEW.Bivalents.DF_IFD$sex <- droplevels(LEW.Bivalents.DF_IFD$sex)

#table(LEW.Bivalents.DF$sex, LEW.Bivalents.DF$hand.foci.count)

@@

lew.ploty_IFD_raw <- ggplot( LEW.Bivalents.DF_IFD,  aes(y=IFD1_ABS, x = sex, na.rm = TRUE))+                     geom_jitter( na.rm = TRUE)+ggtitle("Raw IFD, LEW")+geom_boxplot(aes(x=sex, na.rm = TRUE))
lew.ploty_IFD_raw
#arg, little annoyed that the x NA won't go away

@@

LEW.IFDnorm_scatter.plot <- ggplot(LEW.Bivalents.DF[LEW.Bivalents.DF$hand.foci.count == 2,],  aes(y=IFD1_PER, x = chromosomeLength, color = sex))+  
  scale_color_manual(values=c('#56B4E9','#E69F00'))+
  geom_jitter()+ggtitle("LEW \n Normalized IFD and SC Length")+ylim(c(0,1.1))

@@

                  total.biv = length(Obj.ID),
                  mean.IFD_2CO = mean(IFD1_ABS[which(hand.foci.count == 2)], na.rm = TRUE),
                  mean.IFD.PER = mean(IFD1_PER[which(hand.foci.count == 2)], na.rm = TRUE),
                  n_2CO =  sum(hand.foci.count == 2, na.rm = TRUE) #this is the right syntax
)

@@

@@

PWD.ploty_IFD_raw

PWD.ploty_IFD_PER <- ggplot(PWD.Bivalents.DF[PWD.Bivalents.DF$hand.foci.count == 2,],  aes(y=IFD1_PER, x = sex))+
                          geom_jitter()+ggtitle("Percent IFD, PWD")+geom_boxplot(aes(x=sex))
PWD.ploty_IFD_PER

PWD.IFDnorm_scatter.plot <- ggplot(PWD.Bivalents.DF[PWD.Bivalents.DF$hand.foci.count == 2,],  aes(y=IFD1_PER, x = chromosomeLength, color = sex))+ geom_jitter()+ggtitle("PWD \n Normalized IFD and SC Length")+ylim(c(0,1.1))+scale_color_manual(values=c('#56B4E9','#E69F00'))

@@

@@

```{r KAZ.IFD, echo=FALSE, warning=FALSE}
#make the dataset
#KAZ.Bivalents.DF <- Curated_BivData[Curated_BivData$strain == "KAZ",]

@@

KAZ.IFDnorm_scatter.plot <- ggplot(KAZ.Bivalents.DF[KAZ.Bivalents.DF$hand.foci.count == 2,],  aes(y=IFD1_PER, x = chromosomeLength, color = sex))+ geom_jitter()+ggtitle("KAZ \n Normalized IFD and SC Length")+ylim(c(0,1.1))+scale_color_manual(values=c('#56B4E9','#E69F00'))

@@

@@

```{r MOLF.IFD, echo=FALSE}
#make the dataset
MOLF.Bivalents.DF <- Curated_BivData[Curated_BivData$strain == "MOLF",]
table(MOLF.Bivalents.DF$sex, MOLF.Bivalents.DF$hand.foci.count)

@@

**What is the overlap for cell images in the MLH1 comparisons and these bivalent comparisons?**

should all mice be represented.

@@

@@

```{r Q2.SC.ploy.male, echo=FALSE}
# 1 way anova, is at least 1 strain different from the rest?
# given a lm, how sig is the strain effect

@@

The glm for SC lengths -- sub-grouped by subsp -- indicates that all strain effects are significant for the average mouseisn't met.  SC length.
My prediction for Dom, isn't met.

(indecation that there is more variation for the SC lengths than for gwRR / CO counts)
